---
title: "EDS 223 Week 4 Lab"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}
install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", type = "source")
```

```{r}
library(geodata)
library(terra)
library(tidyverse)
library(tmap)
library(kableExtra)
library(spData)
library(spDataLarge)
```

## Load data

```{r}
# create raster objects
zion_elevation <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
zion_land <- rast(system.file("raster/nlcd.tif", package = "spDataLarge"))

# test class of raster object
class(zion_elevation)

# get summary information 
summary(zion_elevation)
zion_elevation
```

## Make some maps with `tmap`

```{r}
map1 <- tm_shape(zion_elevation) +
  tm_raster(title = "Elevation (m)") +
  tm_layout(legend.outside = TRUE)

map2 <- tm_shape(zion_land) +
  tm_raster(title = "Land cover") +
  tm_layout(legend.outside = TRUE)

tmap_arrange(map1, map2, nrow = 1)
```

## Load more data

```{r}
landsat <- rast(system.file("raster/landsat.tif", package = "spDataLarge"))

nlyr(landsat) # test number of layers in raster object
landsat # see general information
```

```{r}
# create a map showing each of the 4 landsat layers 
tm_shape(landsat) + 
  tm_raster()
```

## Subsetting in a list of rasters

```{r}
landsat3 <- subset(landsat, "landsat_3")

tm_shape(landsat3) + 
  tm_raster()
```

## Global statistics 

```{r}
global(zion_elevation, mean) # take the mean of all pixel values
freq(zion_land) # show each land cover type and their cell count
```

## Indexing datasets/objects 

```{r}
zion_elevation[1, 1] # get value of grid cell in first row & column
zion_elevation[1] # cell value of first cell; same as above

landsat[1] # this produces 4 values; in the first cell, there are 4 values
```

```{r}
# index in a matrix & update its value
zion_elevation [1, 1] <- 0
zion_elevation[1, 1]

landsat[1] <- 0 # this also gets applied to all layers in a list 
landsat[1]
```

```{r}
# if we thought certain values weren't accurate, we could subset to 
# get rid of them 
test_raster <- zion_elevation
test_raster[test_raster < 20] <- NA
```

## Spatial subsetting

#### Using coordinates

```{r}
point <- matrix(c(-113, 37.5), ncol = 2)

# determine elevation value at the point (lat/lon) specified above 
terra::extract(zion_elevation, point) 
```

#### Based on a defined extent 

```{r}
# create a raster with a smaller extent
clip <- rast(xmin = -113.3, xmax = -113, ymin = 37.2, ymax = 37.9,
            resolution = 0.3,
            vals = 1)


# select values that fall within smaller extent
# need to specify 'drop = FALSE' b/c geometry column would get dropped 
zion_elevation_clip <- zion_elevation[clip, drop = FALSE]


# verify that output has fewer values than original
if(ncell(zion_elevation) == nrow(zion_elevation_clip)) {
  warning("clipping did not remove cells")
} else {
  print("clipping removed cells")
}
```

###### Plot to QC it worked

```{r}
tm_shape(zion_elevation) +
  tm_raster()

tm_shape(zion_elevation_clip) +
  tm_raster()

# we can see visually that the clipped raster is smaller
# But we can also check this quantitatively 
if(ext(zion_elevation) == ext(zion_elevation_clip)){
  print("extents match")
} else{
  print("extents do not match")
}
```

#### Masking 

```{r}
# creating our mask 
rmask <- zion_elevation
rmask[rmask < 2000] <- NA # store NA values in all locations w/ elev < 2000

# masking the elevation raster w/ bracket subsetting
masked1 <- zion_elevation[rmask, drop = FALSE]

# masking w/ the mask() function 
masked2 <- terra::mask(zion_elevation, rmask) # raster to mask, mask raster
```

```{r}
 # visual QC-ing
tm_shape(zion_elevation) +
  tm_raster()

tm_shape(masked1) +
  tm_raster()

tm_shape(masked2) +
  tm_raster()
```

## Map Algebra

#### Simple math/logical operations

```{r}
# normal arithmetic 
ex1 <- zion_elevation + zion_elevation # doubles each cells' value
ex2 <- zion_elevation^2 # raises each cells' value to the power of 2

# logical operations
ex3 <- log(zion_elevation) # takes the log of each cells' value
ex4 <- zion_elevation > 5 # determines whether each cell has a value > 5
```

#### Classify w/ matrices

```{r}
# create reclassification matrix
rcl <- matrix(c(1000, 1500, 1, # group 1 ranges from 1000 - 1500 m
                1500, 2000, 2, # group 2 ranges from 1500 - 2000 m
                2000, 2500, 3, # group 3 ranges from 2000 - 2500 m
                2500, 3000, 4), # group 4 ranges from 2500 - 3000 m
                ncol = 3, byrow = TRUE)
rcl

# use reclassification matrix to reclassify elevation raster
reclassified <- terra::classify(zion_elevation, rcl = rcl)

# change reclassified values into factors
values(reclassified) <- as.factor(values(reclassified))


tm_shape(reclassified) + 
  tm_raster()
```

#### Claculating NDVI

```{r}
# create a function to calc NDVI
ndvi_fun <- function(nir, red) {
  (nir - red)/(nir + red)
}

# apply NDVI function to Landsat bands 3 & 4
ndvi_rast <- lapp(landsat[[c(4, 3)]], fun = ndvi_fun)


# map it 
tm_shape(ndvi_rast) + 
  tm_raster(title = "NDVI")
```

#### Focal Opperations 

- We can use the focal() function to perform spatial filtering

- We define the size, shape, and weights of the moving window using a matrix. 

- In the following example we’ll find the minimum value in 9x9 cell neighborhoods.

```{r}
elevation_focal <- focal(zion_elevation, 
                         w = matrix(1, nrow = 9, ncol = 9), # make moving window
                         fun = min) # function to map new values
elevation_focal

tm_shape(elevation_focal) + 
  tm_raster()
```

#### Zonal Operations 

```{r}
# find the mean value in each reclassified groups
zonal(zion_elevation, reclassified, fun = "mean") %>%
  kable(col.names = c("Elevation zone", "Mean elevation (m)")) %>% 
  kable_styling(bootstrap_options = "striped")


# mean elevation across the entire park 
terra::global(zion_elevation, mean)
```

## Geometry Operations

1. Change extent 
2. Change origin 
3. Change resolution 

#### Changing the extent

```{r}
extend <- terra::extend(zion_elevation, c(1, 200)) # add one row and two columns

# will get an error because the extents don't match 
zion_elevation + extend

extend2 <- terra::extend(zion_elevation, extend)

extend + extend2
```

#### Origin 

```{r}
# origin function returns the coordinates of the cell corner closest 
# to the coordinates (0,0).
origin(zion_elevation)
```

#### Changing resolution 

```{r}
# coarsen the resolution of the Zion elevation data by a factor of 5
zion_elevation_coarse <-  aggregate(zion_elevation, 
                                    fact = 5, # scale to aggregate
                                    fun = mean)


# make the resolution finer 
zion_elevation_disagg <- disagg(zion_elevation_coarse, 
                                fact = 5, 
                                method = "bilinear")



# check whether the disaggregated version matches the original
if(identical(zion_elevation, zion_elevation_disagg)){
  print("disaggregated data matches original")
} else {
  warning("disaggregated data does not match original")
}
```

#### Resampling 

```{r}
# create new raster to resample to 
target_raster <- rast(xmin = -113.2, xmax = -112.9, 
                      ymin = 37.14, ymax = 37.5, 
                      nrow = 450, ncol = 460, 
                      crs = crs(zion_elevation))

zion_elevation_resample <- resample(x = zion_elevation, # raster to resample 
                                    y = target_raster,  # raster to resample to 
                                    method = "bilinear")
```
















